---
title: "Identifying Latent Dimensions of Digital Engagement and Social Capital using the 2024 GSS"
subtitle: "Social Survey Analytics Using MCA"
author: "Matthew Beck<br>Min Zhong<br>Kaitlin Kosko<br>Advisor: Dr. Cohen"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---


## Introduction

Our project uses the methodology of Multiple Correspondence Analysis (MCA) in order to perform an exploratory analysis on data derived from the General Social Survey examining the topic of digital inequality.
Dimensionality reduction has been a strategy used by scientists as a means of exploratory data analysis. This allows scientists to visually analyze high-dimensional data that would otherwise be difficult to interpret, thereby recognizing associations between variables. One of the most prominent methods for this is Principal Component Analysis, or PCA, which identifies linear combinations of features which maximize variance. (@gewersPrincipalComponentAnalysis2022, @greenacrePrincipalComponentAnalysis2022)

However, a limitation of PCA is that it only can be used on quantitative, numeric data. For datasets with categorical data, Multiple Component Analysis has been developed. One area in which been able to use MCA is in social scientific fields like sociology, psychology, and political science, especially in the analysis of social surveys which use categorical responses to questions, including yes/no questions and so-called ‘Likert scale’ questions, which ask the respondent to answer in a range from, for example, “strongly disagree” to “strongly agree”. One prominent example is the General Social Survey (GSS), administered by the National Opinion Research Center (NORC) at the University of Chicago, a leading independent social science research organization in the United States. (@GSS) This is the largest such survey in the United States and has been administered for decades to a representative sample of US citizens. The GSS and other such large surveys are perfect for data mining and exploratory data analysis because of their large sample size and the large variety of questions asked on different topics which researchers can use to discover connections between phenomena and underlying patterns of social activity.

In particular, we will be using MCA analysis on the GSS to examine the effects of digital technology on social behavior. In the past decades, digital technology entered into the home, becoming general purpose, woven into everyday life. It has become an integral part of individual life and has changed how people work, study, communicate and participate in society. Because of this important role, it has been discussed through different perspectives. Early studies discussed the “digital divide,” which examines the question of internet access. Recently, research has become broader, expanding to the concept of “digital capital,” a concept which includes digital skills, educational background, and the ability to benefit from online resources (@lybeckDigitalDivideDigital2024). As a result, simple connectivity is no longer sufficient to explain digital inequality; it is a multidimensional concept that involves access, usage patterns, and digital outcomes（@reddickUnderstandingLevelsDigital2024）.

Recent researches suggests that individuals would like to participate in a variety of online activities if they have higher digital capital (@ruiuDigitalCapitalOnline2020). Individuals who shows less online activity usually have lower digital skills, even if they have access to digital technologies. These studies suggest that online participation is associated with broader patterns rather than separate and unrelated behaviors. Meanwhile, Social capital is a complex concept that cannot be measured by a single variable. It has long been discussed by sociologists and it refers to the resources and benefits individuals gain through social networks, community involvement, civic participation, and trust-based relationships. Previous studies have applied principal component analysis(PCA) to identify underlying dimensions of social capital by using multiple survey indicators (@gannonSocialCapitalExploring2020). These approaches suggest that observed measures of social participation and social connectedness may reflect underlying latent dimensions of social capital. 

In this study, we will use data from the 2024 General Social Survey (GSS) to examine whether the association of digital engagement and social capital can be explained through a smaller number of underlying dimensions. Considering that many selected variables are categorical, we apply Multiple Correspondence Analysis (MCA) to identify latent dimensions. The study aims to better understand how these concepts are related in contemporary American society.

## Methodology

In this study, we apply Multiple Correspondence Analysis (MCA) to identify latent dimensions of digital engagement and social capital using categorical data from the 2024 General Social Survey. MCA extends correspondence analysis to settings with multiple categorical variables and can be viewed as a generalization of principal component analysis when variables are categorical rather than quantitative (@Abdi). 

For a dataset, let $X$ denote the indicator matrix. $I$ represents individuals and $J$ represents categories across the variables.  MCA applies correspondence analysis to the $I×J$ indicator matrix $X$, where Z$=X/N$ is the probability matrix, $r$ and $c$ are row and column marginals. Also, $\lambda = \Delta^2$ is the eigenvalue representing the inertia(variance) explained by each dimension. The Chi-square distance between individuals is:

$$ d^2_{i,i'} = \frac{I}{J}\sum^{K}_{k=1}\frac{1}{I_k}\left(x_{ik} - x_{i'k}\right)^2 $$
with the frequency of category $k$, so rare categories carry greater weight. Transition relations link individuals to categories:

$$ F_s(i) = \frac{1}{\sqrt{\lambda_s}}\sum^{J}_{j=1} \sum^{K_j}_{k=1}\frac{x_{ik}}{J}G_s(k) $$
$$ G_s(k) = \frac{1}{\sqrt{\lambda_s}}\sum^{I}_{i=1}\frac{x_{ik}}{I_k}F_s(i) $$

The generalized singular value decomposition (GSVD) applied to the standardized residual matrix in correspondence analysis, then the singular value decomposition (SVD) equation:

$$ \mathbf{D_r}^{1/2}(\mathbf{Z}-\mathbf{rc}^T)\mathbf{D_c}^{-1/2} = \mathbf{P \Delta Q}^T $$
Unlike PCA, which requires quantitative variables and assumes linear relationships, MCA is appropriate here because all GSS variables are categorical. Our study also aligns with Bourdieu's relational approach to social space, where positions are defined by distances and proximities between properties. For those supplementary demographic variables, such as age, education, income, gender), will be projected onto the factorial plane but not contributing to dimension construction. Eigenvalues are corrected by using Benzécri's formula:

$$  c\lambda_l = \left[\left( \frac{K}{K-1} \right)\left( \lambda_l - \frac{1}{K} \right)\right]^2 \quad\text{ if }\quad \lambda_l > \frac{1}{K} $$

Analyses in this study are conducted in RStudio using FactoMineR for computation and Factoshiny for interactive graphical exploration, enabling dynamic selection of variables, adjustment of visualization parameters, and examination of factorial planes. Dimensions are retained based on modified eigenvalue rates above 1/K. Supplementary hierarchical clustering using Ward's method identifies respondent typologies, with interpretation guided by category contributions, squared cosines, and v-tests indicating significant deviations from the mean (@Husson).

#### R libraries used
```{r, message=FALSE}
library('FactoMineR')
library('Factoshiny')
library('factoextra')
library('tidyverse')
library('haven')
library('here')
library('knitr')
library('kableExtra')
```
## Analysis and Results

### Data Exploration and Visualizations
#### Loading dataset
First, the dataset was obtained from the NORC website in an SAS file format. These data were loaded into R and a subset of the full dataset was created, containing only the relevant variables and dropping null observations. This was then exported as a CSV file to be able to be used repeatedly without importing the full dataset.
```{r, eval=FALSE}
#Import entire GSS SAS file
allGSSdata <- read_sas(here::here("..","GSS_sas", "gss7224_r3.sas7bdat"))

#selecting variables to use
variables <- c(
  ## Demographics
  'EDUC', #level of education
  'CLASS', #self-assessed class standing
  'HISPANIC', #id as hispanic
  'WRKSTAT', #employment status
  'MARITAL', #marital status
  'AGE', #age
  'CHILDS', #number of children
  'SEX',
  #'RACE',
  #'INCOME16',
  #'RELIG', #religion
  'RELTRAD', #This is also religion, but a newer, more inclusive. the RELIG variable is a holdeover from earlier surveys
  'ATTEND', #frequency of attending religious services
  'RELPERSN', #strength of religious self-id

  
  ## General Attitudes
  #'TRUST', # Trust in most people
  #'FAIR', #Belief that others will take advantage of them
  
  
  ## Computer usage in core questions - including these results in a lot of missing values
  #'COMPUSE',
  #'WEBMOB',
  #'BROADBAND',
  #'MOBILEDATA',
  
  #Social interaction
  'SOCREL', #frequency of social interaction with relatives
  'SOCOMMUN', #frequency of social interaction with neighbors
  'SOCFREND', #frequency of social interaction with friends
  'SOCBAR', #frequency of going to bar
  
  ## ISSP Digital Societies
  'INTRNETUSE', #frequency of internet use in past 12 months
  'INTPRFSSNL',
  'UNPLUG',
  'INTHOME', #frequency of internet access at home
  'INTPUB', #frewuency of internet access in public
  'INTSKILL', #skill at using internet
  'INTSURF', #skill at online searches
  'INTLITERACY', #knowledge of online information discretion
  'INTAPPS', #skill at learning new software
  'INTCOMM', #frequency of online communication
  'INTSHARE', #frequency of sharing photos and videos
  'INTSEARCH', #frequency of search
  'INTGAME', #uses internet for video games
  'INTSTREAM', #uses internet for streaming music or video
  'INTFNCL', #uses internet for financial transactions
  'INTMEET', #more comfortable meeting people online
  'INTLNLY', #feel lonely without internet
  'INTVIEWS', #frequency of discussing politics online
  'INTRUST', #Trust for people only met online
  'TRPPL' #Trust in others
)


#creating a subset of data including just the selected variables and removing all observations that contain NA values
data <- allGSSdata %>% select(variables) %>% na.omit()

summary(data)

#exporting dataframe to csv file
data %>% write.csv('data.csv')
```

This subset of the data can be easily loaded from the CSV file without taking up the multiple gigabytes of RAM that is used by the full dataset. This subset which will be used for further analysis shows 975 observations of 35 variables.
```{r}
data <- read.csv('data.csv') %>% select(!X)
```

#### Histograms of Variables
Our variables divide into two categories. First, there are variables relating to demographic information of the respondent such as age, sex, and marital status. These will not be used in the main MCA analysis but will be used as supplementary variables. The second category consists of the variables that will be used for MCA analysis, and consist of questions relating the respondent's usage of digital devices as well as questions relating to the respondent's social life.

First, histograms of the supplemental demographic variables are shown below. The variable names are as follows:

| Variable Name |Variable Type      | GSS Description           |
|------------------|--------------------|-----------------------------------------------------|
| AGE               | Numeric (Integer) | Age of Respondent     |
|ATTEND             | Ordinal           | How often Respondent attends religious services|
|CHILDS             | Numeric (Integer) | Number of children|
|CLASS              | Ordinal           | Subjective class identification|
|EDUC               | Numeric (Integer) | Highest year of school completed|
|HISPANIC           | Categorical       | Respondent Hispanic Specific|
|MARITAL            | Categorical       | Marital Status of Respondent|
|RELPERSN           | Ordinal (Likert)  | Respondent consider self a religious person |
|RELTRAD            | Categorical       | Religious tradition |
| SEX               | Categorical       | Respondents sex |
| WRKSTAT           | Categorical       | Labor force status |


```{r, message=FALSE}
#| fig-height: 7

suppVarNames = c("EDUC","CLASS","HISPANIC","WRKSTAT","MARITAL","AGE","CHILDS","SEX","RELTRAD","ATTEND","RELPERSN")

supplementary.long <- data |> pivot_longer(all_of(suppVarNames),names_to = "variable", values_to = "value")

ggplot(supplementary.long, aes(x=value, color=variable,fill=variable)) +
  facet_wrap(~variable, scales='free', ncol=3) +
  geom_bar() +
  theme(legend.position = "none")

```

Next, the histograms of variables relating to usage of and additutes towards digital technology are shown below:

| Variable Name |Variable Type      | GSS Description           |
|------------------|--------------------|-----------------------------------------------------|
| INTAPPS           | Ordinal (Likert)  | Respondent's skill at learning new software |
| INTCOMM           | Ordinal (Likert)  | How often Respondent uses internet for communication|
| INTFNCL           | Ordinal (Likert)  | How often Respondent uses internet for financial transactions|
| INTGAME           | Ordinal (Likert)  | How often Respondent uses internet for video games|
| INTHOME           | Ordinal (Likert)  | Frequency Respondent has home internet access|
| INTLITERACY       | Ordinal (Likert)  | Respondent's knowledge of online information discretion|
| INTLNLY           | Ordinal (Likert)  | Respondent would feel lonely without internet?|
| INTMEET           | Ordinal (Likert)  | Respondent more comfortable meeting people online?|
| INTPRFSSNL        | Ordinal (Likert)  | Frequency Respondent has work/school internet access|
| INTPUB            | Ordinal (Likert)  | Frequency Respondent has internet access in public|
| INTRNETUSE        | Ordinal (Likert)  | Frequency of internet use over past 12 months|
| INTRUST           | Ordinal (Likert)  | How much Respondent trusts people they've only met online|
| INTSEARCH         | Ordinal (Likert)  | How often Respondent uses internet for search|
| INTSHARE          | Ordinal (Likert)  | How often Respondent uses internet for sharing photos/videos|
| INTSKILL          | Ordinal (Likert)  | Respondent's skill at using the internet|
| INTSTREAM         | Ordinal (Likert)  | How often Respondent uses internet for music/video|
|INTSURF            | Ordinal (Likert)  | Respondent's skill at online searches|
| INTVIEWS          | Ordinal (Likert)  | How often Respondent talks about politics online|
| UNPLUG            | Ordinal (Likert)  | Amount Respondent reduced time spent on internet in last 12 months|

```{r, message=FALSE}
#| fig-height: 15

digitalVarNames = c("INTRNETUSE", "INTPRFSSNL", "UNPLUG", "INTHOME", "INTPUB", "INTSKILL", "INTSURF", "INTLITERACY", "INTAPPS", "INTCOMM", "INTSHARE", "INTSEARCH", "INTGAME", "INTSTREAM", "INTFNCL", "INTMEET", "INTLNLY", "INTVIEWS", "INTRUST")
digital.long <- data |> pivot_longer(all_of(digitalVarNames),names_to = "variable", values_to = "value")

ggplot(digital.long, aes(x=value, color=variable,fill=variable)) +
  facet_wrap(~variable, scales='free', ncol=3) +
  geom_bar() +
  theme(legend.position = "none")

```

Finally, below are histograms of variables related to the frequency of respondents attending various social activities as well as a variable that captures the respondent's general trust in other people.

| Variable Name |Variable Type      | GSS Description           |
|------------------|--------------------|-----------------------------------------------------|
| SOCREL            | Ordinal (Likert)  | Frequency of spending evening with relatives |
| SOCBAR            | Ordinal (Likert)  | Frequency of spending evening at bar |
| SOCFREND          | Ordinal (Likert)  | Frequency of spending evening with friends |
| SOCCOMMUN         | Ordinal (Likert)  | Frequency of spending evening with neighbor |
| TRPPL             | Ordinal (Likert)  | Trust in others, 10 point scale |

```{r, message=FALSE}
#| fig-width: 7

socialVarNames = c("SOCREL", "SOCOMMUN","SOCFREND","SOCBAR","TRPPL")
social.long <- data |> pivot_longer(all_of(socialVarNames),names_to = "variable", values_to = "value")

ggplot(social.long, aes(x=value, color=variable,fill=variable)) +
  facet_wrap(~variable, scales='free', ncol=3) +
  geom_bar() +
  theme(legend.position = "none")
```
### Modeling and Results
#### Preparing data for MCA

When running MCA analysis, an $IxJ$ indicator matrix is created where $I$ is the number of observations and $J = \sum^{P}_{p=1}j_p$ with $P$ being the number of variables in the original dataset and $j_p$ being the number of observation levels of variable $p$. This means that $J$ will be large when there are many observation levels of each variable; also, this causes problems when dealing with ordinal data, as the each ordinal level will essentially have an indicator variable created for it during MCA. This can be corrected by transforming the ordinal variables in the original dataset into boolean categorical variables which equal 1 if the response was higher than the median and 0 if less than or equal to the median.

```{R}
#data columns to exclude from MCA
MCAexclude = suppVarNames

#creating dataframe that ONLY has supplementary variables
suppVars = data |> select(all_of(suppVarNames)) |>
  mutate(across(!all_of(c("EDUC","AGE","CHILDS")),factor))      # here I'm making all the columns EXCEPT for EDUC, AGE, and CHILDS into factors because these three are numeric


likert2Boolean <- function(x){
  (x>median(data[[cur_column()]])) |>         #this will be a boolean that will be true if x is greater than the median of the column
    as.integer() |> as.factor()               #then turning boolean into an integer for ease of reading graph and then into a factor
}

MCAdataDigital <- data |>
  select(all_of(digitalVarNames)) |>                        # selecting variables not in the excluded set
  mutate(across(everything(),likert2Boolean))           # applying my likert to boolean function

MCAdata <- data |>
  select(!all_of(MCAexclude)) |>                        # selecting variables not in the excluded set
  mutate(across(everything(),likert2Boolean))      # applying my likert to boolean function

#Create dataframe with both MCA vars and supp vars
MCAdataWithSups <- cbind(MCAdata,suppVars) 

#Creare a vector with the indices of all supplementary variables
suppIndices = length(MCAdata)+1:length(MCAdataWithSups)

#Create a vector with the indices of all quantitative variables (which should also be supplementary)
supQuantInd <-  which(sapply(MCAdataWithSups,class) != 'factor')

#create a vector of all qualitative variables
QualInd <-  which(sapply(MCAdataWithSups,class) == 'factor')
#create a vector of all SUPPLEMENTARY qualitative variables
supQualInd <- QualInd[QualInd %in% suppIndices]
```


#### MCA Analysis 
An MCA analysis is run on all the varaibles relating to sociability and digital engagement. 
```{r}
res.mca <- MCA(MCAdata,graph=FALSE)
res.mca.withsupps <- MCA(MCAdataWithSups,graph=FALSE,quanti.sup = supQuantInd,quali.sup = supQualInd)
fviz_eig(res.mca)
```
As can be seen from the above scree plot, there is one dominant component which explains a relatively high percentage of the total variance, with the remaining components explaining a comparatively much lesser amount of variance.


```{r}
get_mca(res.mca)$contrib |> kable(digits=3) |> kable_styling() |> 
  column_spec(2,color = ifelse(get_mca(res.mca)$contrib[,1] > median(get_mca(res.mca)$contrib[,1]), "red","black")) |>
  column_spec(3,color = ifelse(get_mca(res.mca)$contrib[,2] > median(get_mca(res.mca)$contrib[,2]), "red","black")) |>
  column_spec(4,color = ifelse(get_mca(res.mca)$contrib[,3] > median(get_mca(res.mca)$contrib[,3]), "red","black")) |>
  column_spec(5,color = ifelse(get_mca(res.mca)$contrib[,4] > median(get_mca(res.mca)$contrib[,4]), "red","black")) |>
  column_spec(6,color = ifelse(get_mca(res.mca)$contrib[,5] > median(get_mca(res.mca)$contrib[,5]), "red","black"))
```


When the variables are plotted on the axis representing the principle components, we see clear patterns. In this plot, the variables with "_1" in the name represent the cases where respondents responded in a category higher than the median. Axis 1 appears to capture variation due to overall skill or usage of digital technology while dimension 2 captures variation due to sociability. The direction of these effects must be noted - higher values on Dimension 1 indicate a higher response category, which in turn reflects a lower usage of and skill with digital technologies. A higher vale on dimension 2 correlates with a higher frequency of social events.
```{r}
fviz_mca_var(res.mca, repel=T)
```

The below graph shows the correlation between each variable and the first two components. The same results are visible as above.
```{r}
fviz_mca_var(res.mca, repel=T, choice='var')
```
An interesting result of these graphs is the positioning of some digital engagement variables that appear to correlate more closely with dimension 2 compared to dimension 1. These variables include "INTMEET", "INTGAME", and 'INTTRUST', which are variables which capture some social dimension of digital engagement.

A plot with dimensions 2 and 3 is shown below
```{r}
fviz_mca_var(res.mca, axes=c(2,3),repel=T, choice='var')
```

Plots can be made which include the supplementary variables. The plot shows that most of the supplementary variables are close to the origin, meaning that they have a very small correlation with the two dimesnions. However, 'age' and 'work status' both have a slightly higher correlation with both dimensions when compared with other suppelementary variables.
```{r}
fviz_mca_var(res.mca.withsupps, repel=T, choice='var')
```

Further analysis shows that a higher age is correlated with a higher value on both dimensions; this means higher age is correlated with less digital capital but greater social capital.
```{r}
fviz_mca_var(res.mca.withsupps, repel=T, choice='quanti.sup')
```
```{r}
fviz_mca_var(res.mca.withsupps, repel=T, invisible='var')
```

### Conclusion


## References
