---
title: "Identifying Latent Dimensions of Digital Engagement and Social Capital using the 2024 GSS"
subtitle: "Social Survey Analytics Using MCA"
author: "Matthew Beck<br>Min Zhong<br>Kaitlin Kosko<br>Advisor: Dr. Cohen"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
nocite: |
  @*
---


## Introduction

Our project uses Multiple Correspondence Analysis (MCA) to conduct an exploratory analysis of data from the 2024 General Social Survey (GSS) on the topic of digital inequality. The GSS is administered by the National Opinion Research Center (NORC) at the University of Chicago, a leading independent social science research organization in the United States. (@GSS)

Dimensionality reduction techniques have long been used by researchers as tools for exploratory data analysis. These methods allow researchers to visualize and interpret high-dimensional datasets that would otherwise be difficult to analyze, helping to identify relationships among variables and uncover underlying patterns. One of the most widely used dimensionality reduction techniques is Principal Component Analysis (PCA), which identifies linear combinations of variables that maximize variance within a dataset. (@gewersPrincipalComponentAnalysis2022, @greenacrePrincipalComponentAnalysis2022)

A limitation of PCA, however, is that it can only be applied to quantitative numerical data. To address datasets composed primarily of categorical variables, Multiple Correspondence Analysis (MCA) was developed. MCA has become particularly useful in social science disciplines such as sociology, psychology, and political science, where researchers frequently analyze survey data containing categorical responses. These responses often include binary questions (e.g., yes/no) and Likert-scale questions that ask respondents to indicate their level of agreement or disagreement with a statement, for example in a range from “strongly disagree” to “strongly agree”.

In particular, we will be using MCA analysis on the GSS to examine the effects of digital technology on social behavior. In the past decades, digital technology entered into the home, becoming general purpose, woven into everyday life. It has become an integral part of individual life and has changed how people work, study, communicate and participate in society. Because of this important role, it has been discussed through different perspectives. Early studies discussed the “digital divide,” which examines the question of internet access. The depth of topics covered and the large sample size make the GSS particularly valuable for data mining and exploratory analysis, allowing researchers to identify connections between social phenomena and uncover broader patterns of social behavior.

In this study, MCA is applied to GSS data to examine the relationship between digital engagement and social capital. Over the past several decades, digital technologies have become deeply integrated into everyday life, transforming how individuals work, learn, communicate, and participate in society. As digital technology has become increasingly important, researchers have explored its social consequences from a variety of perspectives. Early studies focused on the concept of the "digital divide," which primarily concerned differences in access to the internet and digital technologies. More recent research has expanded this perspective through the concept of "digital capital," which encompasses not only access but also digital skills, educational background, and the ability to effectively benefit from online resources  (@lybeckDigitalDivideDigital2024). Consequently, digital inequality is now understood as a multidimensional phenomenon involving access, patterns of use, and the outcomes individuals derive from digital participation（@reddickUnderstandingLevelsDigital2024).

Recent studies suggest that individuals with higher levels of digital capital are more likely to engage in a wide variety of online activities (@ruiuDigitalCapitalOnline2020). Conversely, individuals with lower digital skills often participate less in online activities even when they have access to digital technologies. These findings indicate that digital participation reflects broader underlying patterns rather than isolated behaviors. Similarly, social capital is a complex concept that cannot be adequately measured by a single variable. Sociologists generally define social capital as the resources and benefits individuals obtain through social networks, community involvement, civic participation, and relationships based on trust. Previous research has used Principal Component Analysis to identify latent dimensions of social capital by combining multiple survey indicators  (@gannonSocialCapitalExploring2020). These approaches suggest that observed measures of social participation and social connectedness may reflect underlying latent dimensions of social capital. 

In this study, we will use data from the 2024 General Social Survey (GSS) to examine whether the association of digital engagement and social capital can be explained through a smaller number of underlying dimensions. Considering that many selected variables are categorical, we apply Multiple Correspondence Analysis (MCA) to identify latent dimensions. The study aims to better understand how these concepts are related in contemporary American society.  

## Methodology

In this study, we apply Multiple Correspondence Analysis (MCA) to identify latent dimensions of digital engagement and social capital using categorical data from the 2024 General Social Survey. MCA extends correspondence analysis to settings with multiple categorical variables and can be viewed as a generalization of principal component analysis when variables are categorical rather than quantitative ( (@Abdi). 

For a dataset, let $X$ denote the indicator matrix. $I$ represents individuals and $J$ represents categories across the variables.  MCA applies correspondence analysis to the $I×J$ indicator matrix $X$, where Z$=X/N$ is the probability matrix, $r$ and $c$ are row and column marginals. Also, $\lambda = \Delta^2$ is the eigenvalue representing the inertia(variance) explained by each dimension. The Chi-square distance between individuals is:

$$ d^2_{i,i'} = \frac{I}{J}\sum^{K}_{k=1}\frac{1}{I_k}\left(x_{ik} - x_{i'k}\right)^2 $$
with the frequency of category $k$, so rare categories carry greater weight. Transition relations link individuals to categories:

$$ F_s(i) = \frac{1}{\sqrt{\lambda_s}}\sum^{J}_{j=1} \sum^{K_j}_{k=1}\frac{x_{ik}}{J}G_s(k) $$
$$ G_s(k) = \frac{1}{\sqrt{\lambda_s}}\sum^{I}_{i=1}\frac{x_{ik}}{I_k}F_s(i) $$

The generalized singular value decomposition (GSVD) applied to the standardized residual matrix in correspondence analysis, then the singular value decomposition (SVD) equation:

$$ \mathbf{D_r}^{1/2}(\mathbf{Z}-\mathbf{rc}^T)\mathbf{D_c}^{-1/2} = \mathbf{P \Delta Q}^T $$
Unlike PCA, which requires quantitative variables and assumes linear relationships, MCA is appropriate here because all GSS variables are categorical. Our study also aligns with Bourdieu's relational approach to social space, where positions are defined by distances and proximities between properties. For those supplementary demographic variables, such as age, education, income, gender), will be projected onto the factorial plane but not contributing to dimension construction. Eigenvalues are corrected by using Benzécri's formula:

$$  c\lambda_l = \left[\left( \frac{K}{K-1} \right)\left( \lambda_l - \frac{1}{K} \right)\right]^2 \quad\text{ if }\quad \lambda_l > \frac{1}{K} $$

Analyses in this study are conducted in RStudio using FactoMineR for computation and Factoshiny for interactive graphical exploration, enabling dynamic selection of variables, adjustment of visualization parameters, and examination of factorial planes. Dimensions are retained based on modified eigenvalue rates above 1/K. Supplementary hierarchical clustering using Ward's method identifies respondent typologies, with interpretation guided by category contributions, squared cosines, and v-tests indicating significant deviations from the mean (@Husson).

#### R libraries used
```{r, message=FALSE}
library('FactoMineR')
library('Factoshiny')
library('factoextra')
library('tidyverse')
library('haven')
library('here')
library('knitr')
library('kableExtra')
```
## Analysis and Results

### Data Exploration and Visualizations
#### Loading dataset
First, the dataset was obtained from the NORC website in an SAS file format. These data were loaded into R and a subset of the full dataset was created, containing only the relevant variables and dropping null observations. This was then exported as a CSV file to be able to be used repeatedly without importing the full dataset.
```{r, eval=FALSE}
#Import entire GSS SAS file
allGSSdata <- read_sas(here::here("..","GSS_sas", "gss7224_r3.sas7bdat"))

#selecting variables to use
variables <- c(
  ## Demographics
  'EDUC', #level of education
  'CLASS', #self-assessed class standing
  'HISPANIC', #id as hispanic
  'WRKSTAT', #employment status
  'MARITAL', #marital status
  'AGE', #age
  'CHILDS', #number of children
  'SEX',
  #'RACE',
  #'INCOME16',
  #'RELIG', #religion
  'RELTRAD', #This is also religion, but a newer, more inclusive. the RELIG variable is a holdeover from earlier surveys
  'ATTEND', #frequency of attending religious services
  'RELPERSN', #strength of religious self-id
  'RELACTIV', #Frequency of attending religious activites other than service


  
  ## General Attitudes
  #'TRUST', # Trust in most people
  #'FAIR', #Belief that others will take advantage of them
  
  
  ## Computer usage in core questions - including these results in a lot of missing values
  #'COMPUSE',
  #'WEBMOB',
  #'BROADBAND',
  #'MOBILEDATA',
  
  #Social interaction
  'SOCREL', #frequency of social interaction with relatives
  'SOCOMMUN', #frequency of social interaction with neighbors
  'SOCFREND', #frequency of social interaction with friends
  'SOCBAR', #frequency of going to bar
  
  ## ISSP Digital Societies
  'INTRNETUSE', #frequency of internet use in past 12 months
  'INTPRFSSNL',
  'UNPLUG',
  'INTHOME', #frequency of internet access at home
  'INTPUB', #frewuency of internet access in public
  'INTSKILL', #skill at using internet
  'INTSURF', #skill at online searches
  'INTLITERACY', #knowledge of online information discretion
  'INTAPPS', #skill at learning new software
  'INTCOMM', #frequency of online communication
  'INTSHARE', #frequency of sharing photos and videos
  'INTSEARCH', #frequency of search
  'INTGAME', #uses internet for video games
  'INTSTREAM', #uses internet for streaming music or video
  'INTFNCL', #uses internet for financial transactions
  'INTMEET', #more comfortable meeting people online
  'INTLNLY', #feel lonely without internet
  'INTVIEWS', #frequency of discussing politics online
  'INTRUST', #Trust for people only met online
  'TRPPL' #Trust in others
)

#creating a subset of data including just the selected variables and removing all observations that contain NA values
data <- allGSSdata %>% select(variables) %>% na.omit()


#Recoding RACECEN1 to have fewer categories
data <- data |> mutate(RACECEN1.mod = RACECEN1 |> recode_values(
  1 ~ 1, #White
  2 ~ 2, #Black
  3 ~ 3, #American Indian or Alaskan Native
  c(4,5,6,7,8,9,10) ~ 4, # Asian
  c(14,15) ~ 5, # Other
  16 ~ 6 # Hispanic
)) |> select(!RACECEN1)

#exporting dataframe to csv file
data %>% write.csv('data.csv')
```

This subset of the data can be easily loaded from the CSV file without taking up the multiple gigabytes of RAM that is used by the full dataset. This subset which will be used for further analysis shows 975 observations of 35 variables.
```{r}
data <- read.csv('data.csv') %>% select(!X)
```

#### Histograms of Variables
Our variables divide into two categories. First, there are variables relating to demographic information of the respondent such as age, sex, and marital status. These will not be used in the main MCA analysis but will be used as supplementary variables. The second category consists of the variables that will be used for MCA analysis, and consist of questions relating the respondent's usage of digital devices as well as questions relating to the respondent's social life.

#### Demographic Variables (Supplementary Variables)
The demographic profile of our sample ((n = 975)) shows a diverse but educated population. The distribution of education (EDUC) is centered on college-level education, with a median of 14 years of schooling. This indicates that our sample is more educated than the general U.S. population. Age (AGE) is roughly normally distributed with a slight left skew, centered around middle age (median ≈ 50–55). This suggests a mature respondent pool. Marital status (MARITAL) shows a plurality of married individuals, and labor force status (WRKSTAT) indicates that a majority are employed full-time. Interestingly, religious attendance (ATTEND) is highly skewed toward "Never," which matches the broader trend of secularization in American society.

First, histograms of the supplemental demographic variables are shown below. The variable names are as follows:

| Variable Name |Variable Type      | GSS Description           |
|------------------|--------------------|-----------------------------------------------------|
| AGE               | Numeric (Integer) | Age of Respondent     |
|ATTEND             | Ordinal            | How often Respondent attends religious services|
|CHILDS             | Numeric (Integer) | Number of children|
|CLASS              | Ordinal           | Subjective class identification|
|EDUC               | Numeric (Integer) | Highest year of school completed|
|MARITAL            | Categorical       | Marital Status of Respondent|
| RACECEN1.mod          | Categorical       | Race of Respondent (NOTE: recoded from original to have fewer categories)|
|RELPERSN           | Ordinal (Likert)  | Respondent consider self a religious person |
| RELACTIV          | Ordinal           | Frequency of attending religious activities other than services |
|RELTRAD            | Categorical       | Religious tradition |
| SEX               | Categorical       | Respondents sex |
| WRKSTAT           | Categorical       | Labor force status |


```{r, message=FALSE}
#| fig-height: 7

suppVarNames = c("EDUC","CLASS","RACECEN1.mod","WRKSTAT","MARITAL","AGE","CHILDS","SEX","RELTRAD","ATTEND","RELPERSN", "RELACTIV")

supplementary.long <- data |> pivot_longer(all_of(suppVarNames),names_to = "variable", values_to = "value")

ggplot(supplementary.long, aes(x=value, color=variable,fill=variable)) +
  facet_wrap(~variable, scales='free', ncol=3) +
  geom_bar() +
  theme(legend.position = "none")

```

#### Digital Engagement Variables
The digital engagement variables could be separated into three sets based on usage patterns. One set of variables is related to general proficiency and communication, including internet skill (INTSKILL), search ability (INTSURF), online communication (INTCOMM), and frequency of internet use (INTRNETUSE). These variables are heavily skewed toward the higher end of their scales (left-skewed).

In contrast, another set of variables related to specific and non-essential online activities, including gaming (INTGAME) and streaming (INTSTREAM), is skewed toward the lower end (right-skewed). This suggests that while the sample is digitally capable, they do not engage in leisure-oriented online activities. However, financial transactions (INTFNCL) show a more evenly distributed pattern, likely because online banking and shopping have become necessary daily activities rather than optional leisure pursuits.

A third group of variables includes online political discussion (INTVIEWS) and trust in online contacts (INTRUST). These variables show highly skewed distributions toward the lowest categories, suggesting a reluctance to engage in politically charged or high-risk online environments.

Next, the histograms of variables relating to usage of and additutes towards digital technology are shown below:

| Variable Name |Variable Type      | GSS Description           |
|------------------|--------------------|-----------------------------------------------------|
| INTAPPS           | Ordinal (Likert)  | Respondent's skill at learning new software |
| INTCOMM           | Ordinal (Likert)  | How often Respondent uses internet for communication|
| INTFNCL           | Ordinal (Likert)  | How often Respondent uses internet for financial transactions|
| INTGAME           | Ordinal (Likert)  | How often Respondent uses internet for video games|
| INTHOME           | Ordinal (Likert)  | Frequency Respondent has home internet access|
| INTLITERACY       | Ordinal (Likert)  | Respondent's knowledge of online information discretion|
| INTLNLY           | Ordinal (Likert)  | Respondent would feel lonely without internet?|
| INTMEET           | Ordinal (Likert)  | Respondent more comfortable meeting people online?|
| INTPRFSSNL        | Ordinal (Likert)  | Frequency Respondent has work/school internet access|
| INTPUB            | Ordinal (Likert)  | Frequency Respondent has internet access in public|
| INTRNETUSE        | Ordinal (Likert)  | Frequency of internet use over past 12 months|
| INTRUST           | Ordinal (Likert)  | How much Respondent trusts people they've only met online|
| INTSEARCH         | Ordinal (Likert)  | How often Respondent uses internet for search|
| INTSHARE          | Ordinal (Likert)  | How often Respondent uses internet for sharing photos/videos|
| INTSKILL          | Ordinal (Likert)  | Respondent's skill at using the internet|
| INTSTREAM         | Ordinal (Likert)  | How often Respondent uses internet for music/video|
|INTSURF            | Ordinal (Likert)  | Respondent's skill at online searches|
| INTVIEWS          | Ordinal (Likert)  | How often Respondent talks about politics online|
| UNPLUG            | Ordinal (Likert)  | Amount Respondent reduced time spent on internet in last 12 months|

```{r, message=FALSE}
#| fig-height: 15

digitalVarNames = c("INTRNETUSE", "INTPRFSSNL", "UNPLUG", "INTHOME", "INTPUB", "INTSKILL", "INTSURF", "INTLITERACY", "INTAPPS", "INTCOMM", "INTSHARE", "INTSEARCH", "INTGAME", "INTSTREAM", "INTFNCL", "INTMEET", "INTLNLY", "INTVIEWS", "INTRUST")
digital.long <- data |> pivot_longer(all_of(digitalVarNames),names_to = "variable", values_to = "value")

ggplot(digital.long, aes(x=value, color=variable,fill=variable)) +
  facet_wrap(~variable, scales='free', ncol=3) +
  geom_bar() +
  theme(legend.position = "none")

```

#### Social Variables (Active Variables)
The social variables show an informative distribution. Frequency of interaction with relatives (SOCREL) and friends (SOCFREND) is spread across the scale, with a median of "Several Times a Week," indicating active social lives. However, interaction with neighbors (SOCOMMUN) is relatively lower, with a median at "Never" or "Once a Year," suggesting weaker neighborhood-based community ties. Bar attendance (SOCBAR) is similarly low, which may reflect age, lifestyle, or changing social habits. General trust in others (TRPPL) shows a near-normal distribution around the mid-point of the scale (median = 5), indicating moderate levels of social trust in the sample.

Finally, below are histograms of variables related to the frequency of respondents attending various social activities as well as a variable that captures the respondent's general trust in other people.

| Variable Name |Variable Type      | GSS Description           |
|------------------|--------------------|-----------------------------------------------------|
| SOCREL            | Ordinal (Likert)  | Frequency of spending evening with relatives |
| SOCBAR            | Ordinal (Likert)  | Frequency of spending evening at bar |
| SOCFREND          | Ordinal (Likert)  | Frequency of spending evening with friends |
| SOCCOMMUN         | Ordinal (Likert)  | Frequency of spending evening with neighbor |
| TRPPL             | Ordinal (Likert)  | Trust in others, 10 point scale |

```{r, message=FALSE}
#| fig-width: 7

socialVarNames = c("SOCREL", "SOCOMMUN","SOCFREND","SOCBAR","TRPPL")
social.long <- data |> pivot_longer(all_of(socialVarNames),names_to = "variable", values_to = "value")

ggplot(social.long, aes(x=value, color=variable,fill=variable)) +
  facet_wrap(~variable, scales='free', ncol=3) +
  geom_bar() +
  theme(legend.position = "none")
```
### Modeling and Results
#### Preparing data for MCA

When running MCA analysis, an $IxJ$ indicator matrix is created where $I$ is the number of observations and $J = \sum^{P}_{p=1}j_p$ with $P$ being the number of variables in the original dataset and $j_p$ being the number of observation levels of variable $p$. This means that $J$ will be large when there are many observation levels of each variable; also, this causes problems when dealing with ordinal data, as the each ordinal level will essentially have an indicator variable created for it during MCA. This can be corrected by transforming the ordinal variables in the original dataset into boolean categorical variables which equal 1 if the response was higher than the median and 0 if less than or equal to the median.

```{R}
#data columns to exclude from MCA
MCAexclude = suppVarNames

likert2Boolean <- function(x){
  (x>median(data[[cur_column()]])) |>         #this will be a boolean that will be true if x is greater than the median of the column
    as.integer() |> as.factor()               #then turning boolean into an integer for ease of reading graph and then into a factor
}

#creating dataframe that ONLY has supplementary variables
suppVars = data |> select(all_of(suppVarNames)) |>
  mutate(across(all_of(c("ATTEND", "RELPERSN", "RELACTIV")),likert2Boolean)) |>
  mutate(across(!all_of(c("EDUC","AGE","CHILDS")),factor))  # here I'm making all the columns EXCEPT for EDUC, AGE, and CHILDS into factors because these three are numeric


MCAdata <- data |>
  select(!all_of(MCAexclude)) |>                        # selecting variables not in the excluded set
  mutate(across(everything(),likert2Boolean))      # applying my likert to boolean function

#Create dataframe with both MCA vars and supp vars
MCAdataWithSups <- cbind(MCAdata,suppVars) 

#Creare a vector with the indices of all supplementary variables
suppIndices = length(MCAdata)+1:length(MCAdataWithSups)

#Create a vector with the indices of all quantitative variables (which should also be supplementary)
supQuantInd <-  which(sapply(MCAdataWithSups,class) != 'factor')

#create a vector of all qualitative variables
QualInd <-  which(sapply(MCAdataWithSups,class) == 'factor')
#create a vector of all SUPPLEMENTARY qualitative variables
supQualInd <- QualInd[QualInd %in% suppIndices]
```


#### MCA Analysis 
An MCA analysis is run on all the varaibles relating to sociability and digital engagement. 
```{r}
res.mca <- MCA(MCAdata,graph=FALSE)
res.mca.withsupps <- MCA(MCAdataWithSups,graph=FALSE,quanti.sup = supQuantInd,quali.sup = supQualInd)
fviz_eig(res.mca)
```
As can be seen from the above scree plot, there is one dominant component which explains a relatively high percentage of the total variance, with the remaining components explaining a comparatively much lesser amount of variance. The biggest difference between people in this project is related to how they use technology and the internet. The first dimension explains approximately 19% of the total variance, making it the dominant latent dimension, while the second dimension explains about 7.3%. After the first dimension, the percentage of explained variance declines substantially, with subsequent dimensions each accounting for a relatively small proportion of the total variation. This pattern indicates that the primary differences among respondents are captured by the first dimension, which is interpreted as digital capital, reflecting variation in digital technology use and internet engagement. The second dimension captures a smaller but meaningful amount of variation and is interpreted as social capital, representing differences in social interaction and participation. Together, these two dimensions provide the most informative summary of the relationships among the digital engagement and sociability variables used in this study.


```{r}
get_mca(res.mca)$contrib |> kable(digits=3) |> kable_styling() |> 
  column_spec(2,color = ifelse(get_mca(res.mca)$contrib[,1] > 1, "red","black")) |>
  column_spec(3,color = ifelse(get_mca(res.mca)$contrib[,2] > 1, "red","black")) |>
  column_spec(4,color = ifelse(get_mca(res.mca)$contrib[,3] > 1, "red","black")) |>
  column_spec(5,color = ifelse(get_mca(res.mca)$contrib[,4] > 1, "red","black")) |>
  column_spec(6,color = ifelse(get_mca(res.mca)$contrib[,5] > 1, "red","black"))
```


When the variables are plotted on the axis representing the principle components, we see clear patterns. In this plot, the variables with "_1" in the name represent the cases where respondents responded in a category higher than the median. Axis 1 appears to capture variation due to overall skill or usage of digital technology while dimension 2 captures variation due to sociability. The direction of these effects must be noted - higher values on Dimension 1 indicate a higher response category, which in turn reflects a lower usage of and skill with digital technologies. A higher vale on dimension 2 correlates with a lower frequency of social events and generally more pessimistic views on social interaction.

For example, the top contributors are INTSURF_1 (good at online searching) and INTERSKILL_1 (high self rated internet skills), each contributes 7.940% and 7.937% to Dimension 1. The following contributor is INTAPPS_1 (good at learning new apps) at 6.62%. Other important variables include frequent online searching (INTSEARCH_1, 4.73%), online communication (INTCOMM_1, 4.41%), and the ability to judge online information (INTLITERACY_1, 4.17%). Obviously people who score high in Dimension 1 are good with technology, use the internet frequently, and participate in a wide  variety of online activities. Conversely, people who score low are less internet usage, and participate in few online activities. This dimension captures the skill based aspect of digital inequality, indicating that differences in digital ability are the strongest factor in the survey.

In Dimension 2, top contributors are INTRUST_1 (trusting online contacts) at 6.738%, INTMEET_1 (comfortable meeting people online) at 6.396%, and INTRUST_0 (not trusting online contacts) at 6.297%. Followed by SOCFREND_1 (frequent time with friends) at 6.353% and SOCBAR_1 (frequenting bars) at 6.146%. Other important contributors including INTGAME_1 (frequent online gaming) at 4.709%, INTLNLY_1 (feeling lonely without internet) at 2.797%, and SOCOMMUN_1 (frequent interaction with neighbors) at 3.597%. Notably, some variables such like INTRUST_0 and INTRUST_1, INTMEET_0 and INTMEET_1, both contribute to this dimension, meaning that intensity of social attitudes might play a role in defining this dimension. Dimension 2 captures a mix of social behaviors including online (trusting online contacts, meeting people online, gaming) and offline (spending time with friends, going to bars). 

In Dimension 3, the strongest defining variable is INTLNLY_1 (feeling lonely without the internet), contributing 21.102%, followed by INTMEET_1 (comfortable meeting people online) at 15.876%, and SOCOMMUN_1 (frequent interaction with neighbors) at 8.165%. Other important contributors include SOCOMMUN_0 (not interacting with neighbors) at 6.799%, INTLNLY_0 (not feeling lonely without internet) at 6.258%, SOCFREND_1 (frequent time with friends) at 6.563%, INTMEET_0 (not comfortable meeting people online) at 5.926%, and SOCREL_1 (frequent time with relatives) at 4.801%. This dimension captures an aspect social capital which is emotional dependence on the internet. 

Dimension 4 captures basic internet access and usage frequency (INTHOME and INTRNETUSE contributing over 14% and 12%), while Dimension 5 is defined by trust in people and social interaction with relatives (TRPPL and SOCREL contributing over 21% and 13%).

```{r}
fviz_mca_var(res.mca, repel=T)
```

This graph shows that high digital skills and frequent internet use (INTSKILL_1, INTSURF_1, INTAPPS_1, INTSEARCH_1, INTCOMM_1, INTRNETUSE_1, INTSTREAM_1) are all cluster on the right side of the graph, with high positive coordinates on Dimension 1. This confirms that these variables strongly define Digital Capital. Their opposite categories (INTSKILL_0, INTSURF_0) cluster on the left side, representing low digital skills and infrequent use.

Social variables cluster in top (INTMEET_1, SOCFREND_1, SOCBAR_1) and bottom (INTRUST_1, SOCOMMUN_1) sides away from the origin on Dimension 2. Unlike Digital Capital, Social Capital is not a single, simple concept. Different types of social behavior may not always go together. For example, someone may comfortable meeting people online (top) but not trust online contacts (bottom). Or someone may trust online contacts but not be comfortable meeting them.

The below graph shows the correlation between each variable and the first two components. The same results are visible as above.
```{r}
fviz_mca_var(res.mca, repel=T, choice='var')
```
An interesting result of these graphs is the positioning of some digital engagement variables that appear to correlate more closely with dimension 2 compared to dimension 1. These variables include "INTMEET", "INTGAME", and 'INTTRUST', which are variables which capture some social dimension of digital engagement.

A plot with dimensions 2 and 3 is shown below
```{r}
fviz_mca_var(res.mca, axes=c(2,3),repel=T, choice='var')
```

The graph below shows contributions of all active variables to Dimension 1 (explaining 19% of the variance) and Dimension 2 (explaining 7.3% of the variance), with supplementary demographic variables (like age, education, marital status, sex, etc.), and projected onto the same two-dimensional space. So we can see how demographic characteristics relate to Digital Capital (Dimension 1) and Social Capital (Dimension 2). The digital and social categories stay the same positions as in the previous graph: High digital skill variables (INTSKILL_1, INTSURF_1, INTAPPS_1, etc.) cluster on the right side, while social variables such as INTRUST, SOCFREND, and SOCBAR are positioned along Dimension 2. The most of the supplementary variables are close to the origin, meaning that they have a very small correlation with the two dimensions. However, 'age' and 'work status' both have a slightly higher correlation with both dimensions when compared with other supplementary variables.

```{r}
fviz_mca_var(res.mca.withsupps, repel=T, choice='var')
```

Further analysis shows that a higher age is correlated with a higher value on both dimensions; this means higher age is correlated with less digital capital and less social capital. This presents the variable correlation plot from the Multiple Correspondence Analysis (MCA), illustrating the relationships between the study variables and the first two latent dimensions. The first dimension (Dim1) explains 19% of the total variation and is interpreted as digital capital, while the second dimension (Dim2) explains 7.3% and represents social capital. Variables located farther from the origin contribute more strongly to the dimensions. Digital engagement variables, including internet surfing (INTSURF), internet skills (INTSKILL), internet applications (INTAPPS), internet search (INTSEARCH), and internet literacy (INTLITERACY), are strongly associated with Dimension 1, indicating that this dimension reflects higher levels of digital engagement. In contrast, variables such as online trust (INTRUST), comfort meeting people online (INTMEET), online gaming (INTGAME), and socializing with friends (SOCFREND) are more strongly associated with Dimension 2, suggesting that this dimension captures social interaction and relationship building. Notably, several digital variables (INTRUST, INTMEET, and INTGAME) align more closely with the social capital dimension, indicating that some forms of digital engagement are primarily social in nature. Demographic variables generally contribute less to the latent dimensions, although Age shows the strongest association, with additional analyses indicating that older respondents tend to have lower digital capital but higher social capital. Overall, the MCA results suggest that digital capital and social capital are distinct but closely interconnected, with socially oriented digital activities bridging the two dimensions.
```{r}
fviz_mca_var(res.mca.withsupps, repel=T, choice='quanti.sup')
```
The correlation plot shows the relationships between demographic variables and the first two MCA dimensions. Dimension 1 (19%) is most strongly associated with Age, while Children (CHILDS) has a weaker positive association with both dimensions. Education (EDUC) points in the opposite direction of Age along Dimension 1, suggesting that higher educational attainment is associated with lower values on this dimension. Overall, the relatively short vectors indicate that demographic variables explain only a small portion of the variation in the MCA dimensions, with Age contributing the most.
```{r}
fviz_mca_var(res.mca.withsupps, repel=T, invisible='var')
```
The MCA category plot displays the positions of the categorical variable levels across the first two dimensions. Categories that are located close together represent similar response patterns, while categories farther from the origin contribute more strongly to the underlying dimensions. Employment status (WRKSTAT), particularly WRKSTAT_5, is located furthest along Dimension 1, indicating that it has one of the strongest associations with the latent structure. Categories related to religious participation (ATTEND and RELTRAD), marital status (MARITAL), social relationships (RELPERSN), and social class (CLASS) are more clustered around the center, suggesting more moderate contributions to the two dimensions. Overall, the plot indicates that employment status contributes more strongly to the latent dimensions than most of the other demographic categories.

```{r}


fviz_mca_ind(res.mca.withsupps, repel=T, label="none",addEllipses=T,habillage= 27, pointsize=1)+ guides(fill="none") + scale_color_discrete(labels=(c("White","Black","American Indian","Asian","Other","Hispanic")))
```

```{r}

agecats <- data |> mutate(agecat = case_when(
  AGE <= quantile(AGE, 0:4/4)[2] ~ 1,
  AGE <= quantile(AGE, 0:4/4)[3] ~ 2,
  AGE <= quantile(AGE, 0:4/4)[4] ~ 3,
  AGE <= quantile(AGE, 0:4/4)[5] ~ 4
) |> as.factor()
) |> select(agecat)

MCAdata.with.agecats = cbind(MCAdata,agecats)

res.mca.agecats <- MCA(MCAdata.with.agecats,graph=FALSE,quali.sup =25)
fviz_mca_ind(res.mca.agecats, repel=T, label="none",addEllipses=T,habillage= 25, pointsize=1)+ guides(fill="none") + scale_color_discrete(labels=(c("1st Quartile","2nd Quartile", "3rd Quartile", "4th Quartile")))
```

The above two plots show the point cloud of individuals categorized by two of the demographic supplementary variables and with confidence elipses; in the first, they are categorized according to race and in the second by the age quartile. As can be seen, there is poor seperation between the racial groups, with most of the confidence elipses overlapping, though it is clear that there are some differences. As the majority of the survey sample was white, the barycenter of the white demographic group is close to the origin of the plot and so the locations of the other groups can be seen as showing the difference compared to the white group. Overall, Asian-Americans appear to be associated with both moderately higher social and digital capital. The "Other" Category also was notably lower in social capital. When looking at the graph which categorizes individuals by age quartile, a greater degree of separation between categories is apparent, though there is still some overlap. The oldest quartile appears to have the greatsertt difference from the three younger quartiles, though each successive quartile shows a decrease in both social and digital capital.

## Conclusion
This study applied Multiple Correspondence Analysis (MCA) to examine latent dimensions of digital engagement and social capital using categorical data from the 2024 General Social Survey (GSS). Using 975 observations and 35 categorical variables, the analysis explored whether relationships among measures of digital engagement and social connectedness could be explained by a smaller number of underlying dimensions.

The results indicate that two meaningful latent dimensions emerge from the data. The first dimension captures variation in digital technology use and is interpreted as digital capital, reflecting differences in individuals' engagement with digital technologies. The second dimension represents patterns of social interaction and participation and is interpreted as social capital. Together, these dimensions provide a useful framework for understanding how digital engagement and social connectedness are related.

One of the most notable findings is that several technology-related variables—including online gaming (INTGAME), comfort meeting people online (INTMEET), and trust in online contacts (INTRUST)—are more strongly associated with the social capital dimension than with the digital capital dimension. This suggests that digital technologies are not only tools for accessing information but also important platforms for building and maintaining social relationships. The findings therefore indicate that digital capital and social capital are interconnected rather than independent constructs.

Demographic variables generally showed weaker associations with both latent dimensions than measures of digital behavior and social interaction. However, age and employment status were notable exceptions. Older respondents tended to exhibit lower levels of both digital capital and social capital. Similarly, retired individuals were associated with lower scores on both dimensions, while respondents who were currently in school tended to exhibit higher levels of both digital and social capital. These patterns are likely influenced by age, as retired individuals are generally older and students are typically younger.

Several limitations should be considered when interpreting these findings. First, the analytic sample (n = 975) contains a higher proportion of respondents with higher levels of education than the general U.S. population, which may limit the generalizability of the results, particularly for populations that experience greater digital inequality. Second, twenty-four Likert-scale variables were dichotomized into binary variables using median splits to facilitate MCA. Although this approach simplifies interpretation, it also reduces variation in the data and may obscure meaningful differences in respondents' attitudes and behaviors.

Future research should build on these findings by examining whether the relationships between digital capital and social capital remain consistent across different demographic groups, such as urban and rural residents, younger and older adults, or individuals with varying educational backgrounds. Because the GSS provides cross-sectional data, future studies could also incorporate multiple survey waves or longitudinal data to examine how digital engagement and social capital evolve over time and to better assess potential causal relationships. Such research would contribute to a more comprehensive understanding of the role of digital engagement in shaping social inequality and community participation.


## References
